Speed and strength of an epidemic intervention

### 2020-10-19 response.tex

\revtext{In this work the authors propose that ``strength'' and ``speed'' are
complementary frameworks for understanding interventions in infectious
disease models. This is an interesting and appealing idea based on
simple re-arrangements of well-known equations for exponentially growing
epidemics. As the authors themselves note, comparing growth rates r
rather than reproductive numbers R is already well known in ecology.}

\revtext{In my view this paper could bring these ideas to light in infectious
disease more thoroughly and compellingly with an expanded application
section, a section on the relationship to data and/or estimation from
data, improved and expanded section on interventions and the distinct
views provided by (and usefulness of) the two distinct frameworks,
and/or with some extension of the results to population dynamics that
differ from constant rate exponential growth or decay. As it stands it
feels quite minimal, though I appreciate the appeal of the conceptual
point.}

Thank you for this feedback. We agree that these ideas are similar to
ideas that have been discussed in the ecological context, but also feel
that this is a perspective sorely lacking in the disease-modeling world.

We have expanded the
application section with examples from COVID-19, providing a qualitative
examination of how uncertainties in the amount of presymptomatic
transmission can affect conclusions about the efficacy of intervention.
In the beginning of the current pandemic, there was much discussion of
how presymptomatic transmission could make intervention more
difficult---however, some of these conclusions were based on the
assumption that \RR\ is constant, rather than \rr\ (e.g., Hellewell et
al.~2020. ``Feasibility of controlling COVID-19 outbreaks by isolation
of cases and contacts.'' \emph{ Lancet Global Health}). In reality, given
observed \rr, more presymptomatic transmission reduces \RR, meaning that
strength-based interventions will be easier than previously thought. To
our mind, these examples further underline the absence of \rr-based
thinking in the disease-modeling world.

While we appreciate the reviewer's concern that a more data-driven approach could provide further insight, we wanted to keep the focus on the basic general principles without getting bogged down in the details of specific model or data-fitting assumptions.

We now address more explicitly the link between the parameters of the
initial phase and the prospects for long-term control. 
We do not go more deeply here into the subject of non-exponential disease dynamics. 
We are actively working on this topic, and the issues are quite complicated, so we feel it's best to present these in another paper.

@@

``
We note that intervention function $L$ and the strength of intervention $\theta$ need not be calculated explicitly in many contexts: they can usefully be thought of as abstractions of existing modeling practices.
Modelers typically rely on mechanistic models (often based on ordinary differential equations) to model disease spread and evaluate intervention effects.
By doing so, they make implicit assumptions about the shape of $L$ and therefore $\theta$.
''

``Like intervention strength $\theta$, intervention speed $\phi$ is also an abstraction --- that is, the mechanistic models of interventions make implicit assumptions about the shape of the hazard rate $h$ and therefore $\phi$.''

\revtext{
Presumably both frameworks give a decline when the growth rate is
negative and growth when it is positive. Therefore they would never
contradict each other as to whether an intervention was sufficient or
not - is this correct? This should be made explicit.
}

This is correct, and we have worked to be more explicit, specifically at the end of S2 and the end of S3.

### 2020-10-19 responses.tex
ready for meeting?

\jd{I moved this to tex, and then figured out that the only equation here is from the reviewer, which makes it awkward to edit.}

@@

We have expanded the application section with examples from COVID-19. To
our mind, these examples further underline the absence of \rr-based
thinking in the disease-modeling world. We also address more explicitly
the link between the parameters of the initial phase and the prospects
for long-term control. We feel that issues dealing directly with
non-exponential disease dynamics are complicated enough for another
paper.
We have expanded the
application section with examples from COVID-19, providing a qualitative
understanding on how uncertainties in the amount of presymptomatic
transmission can affect conclusions about the efficacy of intervention.
In the beginning of the current pandemic, there was much discussion on
how presymptomatic transmission could make intervention more
difficult---however, some of these conclusions were based on the
assumption that \RR\ is constant, rather than \rr\ (e.g., Hellewell et
al.~2020. ``Feasibility of controlling COVID-19 outbreaks by isolation
of cases and contacts.'' \emph{ Lancet Global Health}. In reality, given
observed \rr, more presymptomatic transmission reduces \RR, meaning that
strength-based interventions will be easier than previously thought. To
our mind, these examples further underline the absence of \rr-based
thinking in the disease-modeling world. While we appreciate the
reviewer's concern that a more data-driven approach could provide
further insight, we are worried that details may take away general ideas
behind speed- and strength-based paradigms. We feel that it is important
to provide a conceptual framework at this stage to a broader audience.
We also address more explicitly the link between the parameters of the
initial phase and the prospects for long-term control. We feel that
issues dealing directly with non-exponential disease dynamics are
complicated enough for another paper.

@@

We have expanded (9) into two steps and attempted to explain the logic
clearly. 
equations?? Or is it OK to leave it like this??

@@

\swp{I like this better for individual vs population-based: The efficacy
of individual-based intervention depends on time since infection (e.g.,
more likely to be isolated during symptomatic stage). The efficacy of
population-based intervention doesn't (e.g., the relative efficacy of
condom doesn't really change). What do you think?}
\jd{This is exactly the kind of detail I'm avoiding at this point. We can get back to it in the next paper?}

>>>>

@@

We have expanded the application section with examples from COVID-19. To
our mind, these examples further underline the absence of \rr-based
thinking in the disease-modeling world. We also address more explicitly
the link between the parameters of the initial phase and the prospects
for long-term control. We feel that issues dealing directly with
non-exponential disease dynamics are complicated enough for another
paper.
\swp{Seems like they're repeated. Delete the first one?}
We have expanded the
application section with examples from COVID-19, providing a qualitative
understanding on how uncertainties in the amount of presymptomatic
transmission can affect conclusions about the efficacy of intervention.
In the beginning of the current pandemic, there was much discussion on
how presymptomatic transmission could make intervention more
difficult---however, some of these conclusions were based on the
assumption that \RR\ is constant, rather than \rr\ (e.g., Hellewell et
al.~2020. ``Feasibility of controlling COVID-19 outbreaks by isolation
of cases and contacts.'' \emph{ Lancet Global Health}. In reality, given
observed \rr, more presymptomatic transmission reduces \RR, meaning that
strength-based interventions will be easier than previously thought. To
our mind, these examples further underline the absence of \rr-based
thinking in the disease-modeling world. While we appreciate the
reviewer's concern that a more data-driven approach could provide
further insight, we are worried that details may take away general ideas
behind speed- and strength-based paradigms. We feel that it is important
to provide a conceptual framework at this stage to a broader audience.
We also address more explicitly the link between the parameters of the
initial phase and the prospects for long-term control. We feel that
issues dealing directly with non-exponential disease dynamics are
complicated enough for another paper.

@@

We have expanded (9) into two steps and attempted to explain the logic
clearly.

@@

### 2020-10-19 speedstrength.tex
ready for meeting?

Here, we argue that the dominance of \RR\ over $r$ in the disease-dynamics literature is excessive: $r$ has been under-rated as a metric. Like \RR,  $r$ can serve as a threshold, and we show that it can also provide a useful metric for difficulty of elimination (cf.~\cite{ferretti2020quantifying}). 
We first generalize the idea that \RR\ measures the difficulty of elimination by showing we can measure an intervention's ``strength'' on the same scale as the reproductive number. 
We then show that we can likewise measure an intervention's ``speed'' on the same scale as the growth rate.
Thus, there is a dual sense in which \rr\ also measures difficulty of elimination. 
We argue that the primacy of $\RR$ over $r$ is partly due to history, and discuss cases where strength provides the better framing for practical disease questions and cases where speed does.
\swp{maybe this is where it might be better to say "speed provides the better framing ..."?}

@@

Since we are interested in invasion and control, we will generally neglect changes in $K(\tau, t)$ through time.
Thus, we will assume $K(\tau, t) \equiv K(\tau)$.
Importantly, this means we are neglecting changes in susceptible proportion through time: that is, assuming that $S(t) \approx S(0)$.
Under this assumption, the renewal equation is equivalent to the Von Foerster equations (see e.g. \cite{fraser2004factors}).
Note that this focus does not mean we are only interested in the initial period of exponential growth: the ability of a disease to increase under conditions characteristic of a naive population is commonly used as a criterion for whether it can be eliminated under general conditions \citep{anderson1985vaccination,fraser2004factors}.
\swp{I feel like there's too much ``something, ...'' phrasing. How about the following?} 
\swp{
Since we are interested in invasion and control, we will generally neglect changes in $K(\tau, t)$ through time and assume $K(\tau, t) \equiv K(\tau)$.
Importantly, this means we are neglecting changes in susceptible proportion through time: $S(t) \approx S(0)$.
Under this assumption, the renewal equation is equivalent to the Von Foerster equations (see e.g. \cite{fraser2004factors}).
However, this focus does not mean we are only interested in the initial period of exponential growth: the ability of a disease to spread under conditions characteristic of a naive population is commonly used as a criterion for whether it can be eliminated under general conditions \citep{anderson1985vaccination,fraser2004factors}.
}
\swp{``conditions characteristic of a naive population'' doesn't really make sense...}

@@

We define the strength of the intervention $L$ to be $\theta = \RR/\Rhat$. 
It is then straightforward to show that $\theta$ is the harmonic mean of $L(\tau)$ weighted by the intrinsic generation-interval distribution:
\begin{equation}
	\theta = 1/\langle 1/L(\tau) \rangle_{g(\tau)}.
	\eqlab{strengthMean}
\end{equation}
Again, we have that the disease cannot spread when $\theta \geq \RR$. 
In other words, $\theta$ generalizes the well-known idea that \RR\ is a metric for how hard a disease is to eliminate: we can measure the ``strength'' of a control measure, this must exceed the strength of the disease (\RR) to achieve elimination.
\swp{The last two/three sentences feel repetitive.}

@@

The disease cannot spread when $\phi \geq r$. In other words, \rr, like \RR, is a metric for how hard a disease is to eliminate: we can measure the ``speed'' of a control measure, this must exceed the speed of the disease (\rr) to achieve elimination.

@@

There has been a great deal on discussion of the importance of pre-symptomatic transmission of COVID-19 \cite{ferretti2020quantifying, he2020temporal, hellewell2020feasibility}. 
Pre-symptomatic transmission is likely to be hard to detect, and therefore hard to prevent. 
Thus, it might be supposed that an increase in the estimated importance of pre-symptomatic transmission would lead to an increase in the estimated difficulty of control.

A generation-interval perspective \cite{park2019practical} can be used to challenge this view. 
More early transmission, for a given value of \swp{observed} \rr, means {shorter generation intervals and therefore} less transmission per individual --- that is, a lower value of \RR. Thus, although earlier transmission could make intervention less effective, it also means that less intervention may be necessary. For interventions like lockdowns, which are generally assumed \swp{to} affect everyone roughly equally, the strength-based perspective thus gives a clear (and somewhat surprising) answer: more early transmission means we will conclude that non-targeted interventions are \emph{more} effective.

@@

We also discussed the question of uncertainties introduced by the unknown magnitude of pre-symptomatic transmission of COVID-19. We showed that the strength-based paradigm provides clear insight into how this uncertainty affects interventions targeted to the general population, while the speed-based paradigm is a clearer way to think about interventions targeted to infected people. We concluded that \swp{``a higher estimate of'' instead of ``increasing''?}increasing pre-symptomatic transmission increases estimated effectiveness of contact-based interventions and decreases estimated effectiveness of symptom-based interventions. In hindsight, these conclusions are consistent with common sense, but in practice the speed-based framework provides a clear way to think about these questions.

@@

Thinking explicitly about the two perspectives can also reduce confusion. Because of the dominance of the strength paradigm, researchers often explore different scenarios while holding \RR\ fixed. Fixed \RR\ is in fact a good default assumption for many endemic diseases. For invading diseases, however, \rr\ is likely to be better constrained by data than \RR. In this case, comparing scenarios while holding \RR\ fixed creates a bias that makes scenarios with faster transmission at the individual level \swp{(i.e., higher proportion of early transmission)} look relatively more dangerous (since these scenarios will have \rr\ faster than the observed value) \citep{eaton2014proportion, powers2014impact, park2019practical}. \swp{I think we can remove the parenthesis in the last sentence}

@@

The importance of speed-based perspectives are still rarely recognized in the case of infectious disease, however. 
Responses to the 2014 Ebola Outbreak in West Africa and the recent COVID-19 outbreak show an over-emphasis on strength at the expense of speed:   
during the early phases of both outbreaks, many disease modelers tried to estimate $\mathcal R_0$ but overlooked $r$.
For example, only 1 out of 7 preliminary analyses of the COVID-19 outbreak that were published as preprints between January 23--26, 2020 reported the doubling time of an epidemic \citep{bedfordncov, imaincov, liuncov, majumderncov, readncov, riouncov, zhaoncov}.
Subsequent studies then relied on strength-based frameworks to evaluate the efficacy of speed-like interventions, such as contact tracing \cite{hellewell2020feasibility,kretzschmar2020impact,kucharski2020contact}, with a few exceptions \cite{ferretti2020quantifying}.
\swp{How is this? Don't want to overwhelm with citations. Seems good enough for me.}
We suggest that infectious disease modelers should be aware of the complementarity of these two frameworks when analyzing disease outbreaks.

>>>>

Here, we argue that the dominance of \RR\ over $r$ in the disease-dynamics literature is excessive: $r$ has been under-rated as a metric. Like \RR,  $r$ can serve as a threshold, and we show that it can also provide a useful metric for difficulty of elimination (cf.~\cite{ferretti2020quantifying}). 
We first generalize the idea that \RR\ measures the difficulty of elimination by showing we can measure an intervention's ``strength'' on the same scale as the reproductive number. 
We then show that we can likewise measure an intervention's ``speed'' on the same scale as the growth rate.
Thus, there is a dual sense in which \rr\ also measures difficulty of elimination. 
We argue that the primacy of $\RR$ over $r$ is partly due to history, and that there are cases where speed provides the better framing for practical disease questions than strength (as well as the reverse). We provide examples of both situations, for both HIV and COVID-19.

@@

Since we are interested in invasion and control, we will generally neglect changes in $K(\tau, t)$ through time.
Thus, we will assume $K(\tau, t) \equiv K(\tau)$.
Importantly, this means we are neglecting changes in susceptible proportion through time: that is, assuming that $S(t) \approx S(0)$.
Under this assumption, the renewal equation is equivalent to the Von Foerster equations (see e.g. \cite{fraser2004factors}).
Note that this focus does not mean we are only interested in the initial period of exponential growth: the ability of a disease to increase under conditions characteristic of a naive population is commonly used as a criterion for whether it can be eliminated under general conditions \citep{anderson1985vaccination,fraser2004factors}.
\swp{I feel like there's too much ``something, ...'' phrasing. How about the following?} 
Since we are interested in invasion and control, we will generally neglect changes in $K(\tau, t)$ through time and assume $K(\tau, t) \equiv K(\tau)$.
Importantly, this means we are neglecting changes in susceptible proportion through time: $S(t) \approx S(0)$.
Under this assumption, the renewal equation is equivalent to the Von Foerster equations (see e.g. \cite{fraser2004factors}).
However, this focus does not mean we are only interested in the initial period of exponential growth: the ability of a disease to spread under conditions characteristic of a naive population is commonly used as a criterion for whether it can be eliminated under general conditions \citep{anderson1985vaccination,fraser2004factors}.
\swp{``conditions characteristic of a naive population'' doesn't really make sense...}

@@

We define the strength of the intervention $L$ to be $\theta = \RR/\Rhat$. 
It is then straightforward to show that $\theta$ is the harmonic mean of $L(\tau)$ weighted by the intrinsic generation-interval distribution:
\begin{equation}
	\theta = 1/\langle 1/L(\tau) \rangle_{g(\tau)}.
	\eqlab{strengthMean}
\end{equation}
As in the constant-$L$ case above, we have that the disease cannot spread when $\theta \geq \RR$. 
In other words, $\theta$ generalizes the well-known idea that \RR\ is a metric for how hard a disease is to eliminate: we can measure the ``strength'' of a control measure, this must exceed the strength of the disease (\RR) to achieve elimination.
\swp{The last two/three sentences feel repetitive.}

@@

The disease cannot spread when $\phi \geq r$. In other words, \rr, like \RR, is a metric for how hard a disease is to eliminate: we can measure the ``speed'' of a control measure, this must exceed the speed of the disease (\rr) to achieve elimination. Since both $\phi \geq r$ and $\theta \geq \RR$ are conditions for control, they are necessarily equivalent: the speed paradigm does not provide a different answer, it provides a different way of thinking about the steps to the correct answer.

@@

There has been a great deal of discussion of the importance of pre-symptomatic transmission of COVID-19 \cite{ferretti2020quantifying, he2020temporal, hellewell2020feasibility}. 
Pre-symptomatic transmission is likely to be hard to detect, and therefore hard to prevent. 
Thus, it might be supposed that an increase in the estimated importance of pre-symptomatic transmission would lead to an increase in the estimated difficulty of control.

A generation-interval perspective \cite{park2019practical} can be used to challenge this view. 
More early transmission, for a given value of observed \rr, means {shorter generation intervals and therefore} less transmission per individual --- that is, a lower value of \RR. Thus, although earlier transmission could make intervention less effective, it also means that less intervention may be necessary. For interventions like lockdowns, which are generally assumed to affect everyone roughly equally, the strength-based perspective thus gives a clear (and somewhat surprising) answer: more early transmission means we will conclude that non-targeted interventions are \emph{more} effective.

@@

We also discussed the question of uncertainties introduced by the unknown magnitude of pre-symptomatic transmission of COVID-19. We showed that the strength-based paradigm provides clear insight into how this uncertainty affects interventions targeted to the general population, while the speed-based paradigm is a clearer way to think about interventions targeted to infected people. We concluded that a higher estimate of pre-symptomatic transmission increases estimated effectiveness of contact-based interventions and decreases estimated effectiveness of symptom-based interventions. In hindsight, these conclusions are consistent with common sense, but in practice the speed-based framework provides a clear way to think about these questions.

@@

Thinking explicitly about the two perspectives can also reduce confusion. Because of the dominance of the strength paradigm, researchers often explore different scenarios while holding \RR\ fixed. Fixed \RR\ is in fact a good default assumption for many endemic diseases. For invading diseases, however, \rr\ is likely to be better constrained by data than \RR. In this case, comparing scenarios while holding \RR\ fixed creates a bias that makes scenarios with faster transmission at the individual level (i.e., higher proportion of early transmission) look relatively more dangerous, because these scenarios will have \rr\ faster than the observed value \citep{eaton2014proportion, powers2014impact, park2019practical}.

@@

The importance of speed-based perspectives are still rarely recognized in the case of infectious disease, however. 
Responses to the 2014 Ebola Outbreak in West Africa and the recent COVID-19 outbreak show an over-emphasis on strength at the expense of speed:   
during the early phases of both outbreaks, many disease modelers tried to estimate $\mathcal R_0$ but overlooked $r$.
For example, only 1 out of 7 preliminary analyses of the COVID-19 outbreak that were published as preprints between January 23--26, 2020 reported the doubling time of an epidemic \citep{bedfordncov, imaincov, liuncov, majumderncov, readncov, riouncov, zhaoncov}.
Subsequent studies then relied on strength-based frameworks to evaluate the efficacy of speed-like interventions, such as contact tracing \cite{hellewell2020feasibility,kretzschmar2020impact,kucharski2020contact}, with a few exceptions \cite{ferretti2020quantifying}.
We suggest that infectious disease modelers should be aware of the complementarity of these two frameworks when analyzing disease outbreaks.
